Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-15 01-introduction.qmd
abstract , intro, antecedentes JC

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023]. Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021]. Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023]. In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain. We examine this association in the context of COVID 19 in Chile, a period characterized by several challenges regarding the political and economic system that could have affected the way in which the role of individual performance and social outcomes are deemed.

Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022]. This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008]. Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile during the COVID pandemic, and to what extent this phenomenon is associated with the perception of meritocracy and its variations.

@@

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021]. A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they would be the most affected in critical times.

Against this backdrop, Chile serves as an example among Latin American countries due to its strong sanitary measures. These measures significantly impacted the labor market, leading to important consequences for unemployment and income distribution. It is importantto approach any claims about the influence of the COVID-19 pandemic with caution. Our findings are situated within a broader context of increasing socioeconomic risks and vulnerabilities, which were heightened by the sanitary restrictions implemented during the pandemic.

>>>>

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023]. Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021]. Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023]. In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain. We examine this association in the context of COVID 19 in Chile, a period characterized by several challenges regarding the political and economic system that could have affected the way in which the role of individual performance and social outcomes are deemed.

@@

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021]. A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008]. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals.

### 2025-01-15 02-antecedentes.qmd
abstract , intro, antecedentes JC

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items has been diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects: the ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

@@

The relationship between performance and outputs brings about to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find related research which incorporates the support of different distributive principles with regard to education. For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. However, the relationship between meritocracy and the justification of inequality within the educational domain remains underexplored. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.

@@

_This study_

The present research aims at identifying changes in the justification of educational inequalities in Chile, with a particular focus in its relationship with meritocratic perceptions. The time frame for this analysis covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile). Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide. Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval.

As there are measures before and after the pandemic period, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].

Within this framework, we point out that periods of rising socioeconomic and health vulnerabilities could weaken the justification of social inequalities. This change would be at least in part linked to a challenge of the meritocratic ideals. Chile provides a significant case for studying inequality, as it exemplifies a prosperous yet highly unequal society within Latin America. Under such conditions, a shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions. Chile implemented sanitary controls early in the pandemic, initiating lockdowns on March 18, 2020, and lifting all restrictions by March 5, 2023. In this context, Chile stands out as one of the strictest countries globally and within South America in terms of pandemic-related policies. According to the Stringency Index [@hale_global_2021]—ranging from 0 to 100 (100 being the strictest) and encompassing measures like school closures, workplace shutdowns, and travel bans—the global median was 42.54, while Chile scored 58.1. This placed Chile 36th among 184 countries and 4th in South America, following Peru (60.1), Suriname (60.5), and Venezuela (80.5).

Within the period analyzed (2016-2022), an important cycle of protests known as the "social outburst" took place between October 2019 and March 2020 [@somma_no_2021]. However, the protest drastically decreased as a result of the lockdown [@joignant_informe_2020]. In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiograficambio_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic. 
Hereby, a relevant aspect is the transformation of the labor market and its impact on income distribution [@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,]. In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6. In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade. By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.

Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:

@@

Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019]. Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps. However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification. In line with this evidence, we expect that:

>>>>

In the area of survey studies, the operationalization of educational inequality justification in terms of questionnaire items has been diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects: the ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs. Although such items covering inequality justification are present in emblematic comparative studies, still there are few studies focusing specifically in the justification of educational inequality.

@@

Whether inequalities in education are considered just or not lead us almost naturally to the concepts of merit and meritocracy. Meritocracy, a concept coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to criticize a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often considered inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find research which incorporates the support of different distributive principles with regard to education. For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003] is linked to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.

@@

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.

_The COVID pandemic and public concerns about inequality_

The time frame for this study covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile). Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide. However, and as several studies have shown, this period appeared to trigger attitudinal changes that are relevant to explore, both in themselves and as antecedent to future studies. In this sense, in this paper we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].

A shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions[@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,]. In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6. In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade. By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.

Chile implemented sanitary controls early in the pandemic, initiating lockdowns on March 18, 2020, and lifting all restrictions by March 5, 2023. In this context, Chile stands out as one of the strictest countries globally and within South America in terms of pandemic-related policies. According to the Stringency Index [@hale_global_2021]—ranging from 0 to 100 (100 being the strictest) and encompassing measures like school closures, workplace shutdowns, and travel bans—the global median was 42.54, while Chile scored 58.1. This placed Chile 36th among 184 countries and 4th in South America, following Peru (60.1), Suriname (60.5), and Venezuela (80.5). Besides, an important cycle of protests known as the "social outburst" took place between October 2019 and March 2020 [@somma_no_2021]. As a consequence there were two constitutional processes happening in this very same period after the massive social upheavals. However, the protest drastically decreased as a result of the lockdown [@joignant_informe_2020].

In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiograficambio_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:

@@

Our second argument is based on the evidence that relates rational interests to justifications and preferences [@sonhing_failure_2019]. Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about — and more supportive of proposals to close — wealth-based achievement gaps than Black-White or Hispanic-White gaps. In line with this evidence, we expect that:

### 2025-01-15 paper.qmd
abstract , intro, antecedentes JC

---
title: Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: 
      - top=30mm
      - left=20mm
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
affiliations: 
  - id: 1
    name: Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. To what extent are such educational disparities justified in society? And, what are the main factors driving this kind of justification? This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy. The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.

>>>>

---
title: Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: 
      - top=30mm
      - left=20mm
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
affiliations: 
  - id: 1
    name: Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. This paper is guided by the following research questions: To what extent are such educational disparities justified in society? Is the justification of educational inequality affected in periods of personal and social vulnerability, as in the COVID pandemic? And, what are the main factors driving this kind of justification? The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. This economic and cultural environment has promoted meritocratic ideals, whereby individual talent and effort are considered key to get ahead in life, disregarding opportunities linked to social origin. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic). Furthermore, we argue that such changes could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2023, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.
